---
title: "Population trends"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(cwi)
library(camiller)
```

6/22: The other notebook was getting unwieldy, so splitting these into smaller notebooks. While I work on it, stuff is just copy/pasted from the other notebook to sit here until I sort it into its new notebook.

* Change by race *(major three/four and other?)* - state, county; town?
* Change by household type/size *(see town reports)* - state, county; town?
* Change by income *(need to define bands)* - state, county; town?

```{r}


acs_years <- list("2011" = 2011, "2012" = 2012, "2013" = 2013, "2014" = 2014, "2015" = 2015, "2016" = 2016, "2017" = 2017, "2018" = 2018)

tenure <- acs_years %>% map(~multi_geo_acs(table = "B25003", year = ., new_england = F))
tenure_bind <- Reduce(rbind, tenure) %>% label_acs()

hh_type <- acs_years %>% map(~multi_geo_acs(table = "B11001", year = ., new_england = F))
hh_type_bind <- Reduce(rbind, hh_type) %>% label_acs()

```