---
title: "Population trends"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(cwi)
library(camiller)
```

6/22: The other notebook was getting unwieldy, so splitting these into smaller notebooks. While I work on it, stuff is just copy/pasted from the other notebook to sit here until I sort it into its new notebook.

* Change by age *(need to define bands)* - state, county; town?
* Change by race *(major three/four and other?)* - state, county; town?
* Change by household type/size *(see town reports)* - state, county; town?
* Change by income *(need to define bands)* - state, county; town?

```{r}
out <- list()

acs_years <- list("2011" = 2011, "2012" = 2012, "2013" = 2013, "2014" = 2014, "2015" = 2015, "2016" = 2016, "2017" = 2017, "2018" = 2018)

#median income?? - b19013
#pop by race - b03002
#tenure - b25003
#hh type - b11001

race <- acs_years %>% map(~multi_geo_acs(table = "B03002", year = ., new_england = F))
race_bind <- Reduce(rbind, race) %>% label_acs()

tenure <- acs_years %>% map(~multi_geo_acs(table = "B25003", year = ., new_england = F))
tenure_bind <- Reduce(rbind, tenure) %>% label_acs()

hh_type <- acs_years %>% map(~multi_geo_acs(table = "B11001", year = ., new_england = F))
hh_type_bind <- Reduce(rbind, hh_type) %>% label_acs()

out$race_state_county_town_2011_2018 <- race_bind
out$tenure_state_county_town_2011_2018 <- tenure_bind
out$hh_type_state_county_town_2011_2019 <- hh_type_bind
```




Pop by age, this will need total pops table
```{r}
pop_by_age <- pop_tables$sex_by_age_state_county_town_2011_2018 %>%
	separate(label, into = c("total", "gender", "age"), sep = "!!", fill = "right") %>% 
	clean_names() %>%
	filter(!grepl("_001|_002|_026", variable)) %>% #remove total, total male, total female
  select(-gender, -total, -variable, -state) %>% 
	mutate(age = str_replace_all(age, " ", "_") %>% 
				 	str_to_lower()) %>% 
	mutate(age = as.factor(age) %>% 
				 	fct_relevel(., "under_5_years", "5_to_9_years") %>% 
				 	fct_collapse(.,
				 							 `15_to_19_years` = c("15_to_17_years", "18_and_19_years"),
				 							 `20_to_24_years` = c("20_years", "21_years", "22_to_24_years"),
				 							 `60_to_64_years` = c("60_and_61_years", "62_to_64_years"),
				 							 `65_to_69_years` = c("65_and_66_years", "67_to_69_years"))) %>% 
	group_by(year, level, geoid, name, county, var = age) %>% 
	summarise(estimate = sum(estimate),
						moe = moe_sum(moe = moe, estimate = estimate)) %>% 
	bind_rows(total_pop) %>%
	group_by(year, level, geoid, name, county) %>% 
	mutate(moe = replace_na(moe, 0),
				 moe = round(moe, 0)) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate, moe = moe)

age_lut <- tibble(agegrp = c(99, seq(1:18)),
									age = unique(pop_by_age$var)) %>% 
	add_row(agegrp = 0, age = "under_5_years")

period_lut <- tibble(
	estimate_date = c(
		"April 2000 base estimate",
		"July 1, 2000 estimate",
		"July 1, 2001 estimate",
		"July 1, 2002 estimate",
		"July 1, 2003 estimate",
		"July 1, 2004 estimate",
		"July 1, 2005 estimate",
		"July 1, 2006 estimate",
		"July 1, 2007 estimate",
		"July 1, 2008 estimate",
		"July 1, 2009 estimate",
		"2010 census count",
		"July 1, 2010 estimate"),
	year = seq(1:13))
```

```{r}
int_pop <- intercensal %>% 
	mutate(geoid = paste(state, county, sep = "")) %>% 
  select(geoid, name = ctyname, year, agegrp, estimate = tot_pop) %>% 
	left_join(age_lut, by = "agegrp") %>% 
	left_join(period_lut, by = "year") %>% 
	select(-year, -agegrp) %>% 
	rename(var = age, year = estimate_date) %>% 
	mutate(level = "2_county", county = NA, moe = 0,
				 var = as.factor(var) %>% 
				 	fct_relevel(., "under_5_years", "5_to_9_years")) %>% 
	group_by(year, level, geoid, name, county, var) %>% 
	summarise(estimate = sum(estimate), moe = sum(moe)) %>% 
	ungroup() %>% 
	group_by(year, level, geoid, name, county) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate, moe = moe)
```