---
title: "Commute times by tenure"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(ipumsr)
library(srvyr)
```

```{r gg etc}
theme_set(hrbrthemes::theme_ipsum_rc(base_family = "Lato Regular"))

#urban colors
pal <- c("#1696d2", "#fdbf11", "#d2d2d2", "#ec008b", "#55b748")
```

I'm not sure if this makes the most sense or not, but commute times here are given as share by time range (mean is influenced by some very high and low commute times) by county and the state, then by tenure.

LODES data are an alternative but a little dated (coupled with the fact that most high income workers who often had longer commutes are now working from home). Maybe discuss this with the team when it comes to the write up.

These are top-coded (each survey uses a different topcode, but all are over 120 mins).

The ranges I'm using are:

* 1 to 15 minutes - short commute
* 16 to 30 minutes - average range for CT and US
* 31 to 60 minutes - average range for transit riders
* 61 to 120 minutes - long commute
* More than 120 minutes - very long commute

## Setup PUMS file

```{r}
names <- tibble(countyfip = seq(from = 1, to = 15, by = 2),
								name = c("Fairfield County", "Hartford County", "Litchfield County", "Middlesex County", "New Haven County", "New London County", "Tolland County", "Windham County"))

ddi <- read_ipums_ddi("../input_data/usa_00049.xml")

pums <- read_ipums_micro(ddi, verbose = F) %>% 
	mutate_at(vars(YEAR, OWNERSHP), as_factor) %>% 
	mutate_at(vars(HHINCOME, PERWT), as.numeric) %>% 
	janitor::clean_names() %>% 
	left_join(names, by = "countyfip") %>% 
	mutate(tenure = if_else(ownershp == "Rented", "Renter", "Owner")) %>%
	mutate(trantime_f = as.factor(trantime),
				 trantime_n = as.numeric(trantime_f)) %>%
	mutate(commute_time = cut(trantime_n, breaks = c(0, 15, 30, 60, 120, Inf), labels = c("Up to 15 mins", "16 to 30 mins", "31 to 60 mins", "61 to 120 mins", "More than 120 mins"), right = T)) %>%
	select(year, pernum, perwt, name, ownershp, tenure, hhincome, trantime, commute_time)
```

## By county and state, no disaggregation.

```{r}
des <- pums %>%
	filter(hhincome != 9999999, ownershp != "N/A", trantime != 0) %>% 
	as_survey_design(., ids = 1, wt = perwt)

total_county <- des %>% 
	select(name, year, perwt) %>% 
	mutate(commute_time = "Total") %>% 
	group_by(name, year, commute_time) %>% 
	summarise(commuters = survey_total(perwt)) %>% 
	select(-commuters_se)

total_state <- total_county %>% 
	mutate(name = "Connecticut") %>% 
	ungroup() %>% 
	group_by(name, year, commute_time) %>% 
	summarise(commuters = sum(commuters))

county_com <- des %>%
	select(name, year, perwt, commute_time) %>% 
	group_by(name, year, commute_time) %>% 
	summarise(commuters = survey_total(perwt)) %>%
	select(-commuters_se)

state_com <- county_com %>% 
	mutate(name = "Connecticut") %>% 
	ungroup() %>% 
	group_by(name, year, commute_time) %>% 
	summarise(commuters = sum(commuters))

commute <- bind_rows(state_com, total_state, county_com, total_county) %>% 
	mutate(year = as.factor(year),
				 commute_time = as.factor(commute_time) %>% 
				 	fct_relevel(., "Up to 15 mins") %>% 
				 	fct_relevel(., "Total", after = Inf)) %>% 
	arrange(name, year, commute_time) %>% 
	group_by(name, year) %>% 
	camiller::calc_shares(group = commute_time, denom = "Total", value = commuters) %>% 
	ungroup()

write_csv(commute, file = "../output_data/commute_time_2000_2018.csv")
```

```{r fig.height=8, fig.width=8}
commute %>% 
	filter(!is.na(share),
				 year == "2018") %>% 
	mutate(commute_time = fct_rev(commute_time)) %>% 
	ggplot(aes(share, commute_time)) +
	geom_col(width = .7, position = position_dodge(.8), fill=pal[1]) +
	geom_text(aes(label = scales::percent(share, accuracy = 1)), position = position_dodge(.8), hjust = -.1, family = "Lato Regular", size = 3) +
	scale_x_continuous(expand = expansion(mult = c(.05, .272))) +
	facet_wrap(facets = "name") + 
	labs(title = "Share of Commuters by Travel Time to Work, 2018", x = "", y = "") +
	theme(plot.title.position = "plot",
				plot.title = element_text(family = "Lato Bold", size = 11),
				panel.grid.major = element_blank(), 
				panel.grid.minor = element_blank(),
				legend.position = "none",
				axis.text.x = element_blank(),
				axis.text.y = element_text(colour = "black", family = "Lato Regular", size = 9),
				strip.text.x = element_text(hjust= .5, colour = "black", family = "Lato Regular", size = 9))

ggsave(filename = "../output_data/corrected_charts/commuters.png", dpi = 300, width = 6.5)
ggsave(filename = "../output_data/corrected_charts/commuters.svg", dpi = 300, width = 6.5)
```

I'm only showing 2018 here even though 2010 and 2000 are listed in the csv (not much change over time...). It's still a very long table, but more analysis follows.

```{r}
commute %>% 
	filter(year == "2018") %>% 
	kableExtra::kable()
```

## By tenure

To do this, I need to make a LUT of owner and renter households

```{r eval = F}
total_county_tenure <- des %>% 
	select(name, year, tenure, perwt) %>% 
	mutate(commute_time = "Total") %>% 
	group_by(name, year, tenure, commute_time) %>% 
	summarise(commuters = survey_total(perwt)) %>% 
	select(-commuters_se)

total_state_tenure <- total_county_tenure %>% 
	mutate(name = "Connecticut") %>% 
	ungroup() %>% 
	group_by(name, year, tenure, commute_time) %>% 
	summarise(commuters = sum(commuters))

county_com_tenure <- des %>%
	select(name, year, perwt, tenure, commute_time) %>% 
	group_by(name, year, tenure, commute_time) %>% 
	summarise(commuters = survey_total(perwt)) %>%
	select(-commuters_se)

state_com_tenure <- county_com_tenure %>% 
	mutate(name = "Connecticut") %>% 
	ungroup() %>% 
	group_by(name, year, tenure, commute_time) %>% 
	summarise(commuters = sum(commuters))

commute_tenure <- bind_rows(state_com_tenure, total_state_tenure, county_com_tenure, total_county_tenure) %>% 
	mutate(year = as.factor(year),
				 tenuer = as.factor(tenure),
				 commute_time = as.factor(commute_time) %>% 
				 	fct_relevel(., "Up to 15 mins") %>% 
				 	fct_relevel(., "Total", after = Inf)) %>% 
	arrange(name, year, tenure, commute_time) %>% 
	group_by(name, year, tenure) %>% 
	camiller::calc_shares(group = commute_time, denom = "Total", value = commuters) %>% 
	ungroup()

write_csv(commute_tenure, file = "../output_data/commute_time_by_tenure_2000_2018.csv")
```