---
title: "Cost burden"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(ipumsr)
library(tidyverse)
library(srvyr)
```

Looking at cost burden by a few demographic factors and geographic levels. Separately, looking at cost burden by the presence of a household member with a disability.

```{r}
ddi <- read_ipums_ddi("../input_data/usa_00036.xml")
read_pums <- read_ipums_micro(ddi, verbose = F)

#not currently doing anything with age
ct_pums <- read_pums %>% 
	mutate_at(vars(YEAR, PUMA, OWNERSHP, OWNERSHPD, SEX, RACE, RACED, HISPAN, HISPAND,
								 DIFFREM, DIFFPHYS, DIFFMOB, DIFFCARE, DIFFEYE, DIFFHEAR), as_factor) %>% 
	mutate_at(vars(PERWT, HHWT), as.numeric) %>% 
	mutate_at(vars(HHINCOME, POVERTY, AGE, OWNCOST, RENTGRS), as.integer) %>% 
	#below flags for 0-1x, 1-2x, 2x+ fpl, but could use other cut points
	mutate(fpl = if_else(POVERTY < 100, "0 to 100% FPL", "201% FPL or more")) %>% 
	mutate(fpl = if_else((POVERTY >= 100 & POVERTY <=200), "101% to  200% FPL", fpl)) %>% 
	# Asian is perhaps the only other race category that might not be too small
	mutate(race2 = if_else(HISPAN == "Not Hispanic", as.character(RACE), "Latino")) %>% 
	mutate(race2 = as.factor(race2) %>% 
				 	fct_recode(Black = "Black/African American/Negro") %>%
				 	fct_other(keep = c("White", "Black", "Latino"), other_level = "Other race") %>%
				 	fct_relevel("White", "Black", "Latino", "Other race")) %>% 
	janitor::clean_names()

#logical flag for any disability
ct_pums$has_disability <- apply(ct_pums, 1, function(x) any(grep("Has|Yes", x)))
```

## State-level

Cost burden by each and combinations of

* race/ethnicity
* gender
* tenure
* percent FPL

```{r}
state_out <- list()

hhdes <- ct_pums %>%
	group_by(year) %>% 
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>% 
	as_survey_design(., ids = 1, wt = hhwt)

count_hhlds <- hhdes %>%
	select(year, hhwt, sex, race2, ownershp, fpl) %>% 
	group_by(year, sex, ownershp, race2, fpl) %>% 
	summarise(total_hhlds = survey_total(hhwt)) %>% 
	mutate(region = "Connecticut") %>% 
	select(region, everything())

state_out$count_hhlds <- count_hhlds

rent_burd <- hhdes %>% 
	filter(ownershp == "Rented") %>% 
	select(year, race2, sex, ownershp, fpl, rentgrs, hhincome, hhwt) %>% 
	mutate(pct_income_to_housing = (rentgrs * 12) / hhincome,
				 is_burdened = if_else(pct_income_to_housing > .3, T, F)) %>% 
	#capture hhlds with negative income as burdened
	mutate(is_burdened = if_else(pct_income_to_housing < 0, T, is_burdened)) %>% 
	#some people have no rent and no income. filtering them out
	filter(!is.na(is_burdened)) %>%  
	group_by(year, race2, sex, is_burdened, ownershp, fpl) %>% 
	summarise(hhlds = survey_total(hhwt)) %>% 
	left_join(count_hhlds, by = c("year", "race2", "sex", "ownershp", "fpl")) %>% 
	mutate(share = round(hhlds / total_hhlds, 3)) %>% 
	select(-contains("total")) %>% 
	select(region, everything())

own_burd <- hhdes %>% 
	filter(ownershp == "Owned or being bought (loan)") %>% 
	select(year, race2, sex, ownershp, fpl, owncost, hhincome, hhwt) %>% 
	mutate(pct_income_to_housing = (owncost * 12) / hhincome,
				 is_burdened = if_else(pct_income_to_housing > .3, T, F)) %>% 
	#capture hhlds with negative income as burdened
	mutate(is_burdened = if_else(pct_income_to_housing < 0, T, is_burdened)) %>% 
	#filter no housing costs/no income people
	filter(!is.na(is_burdened)) %>%  
	group_by(year, race2, sex, is_burdened, ownershp, fpl) %>% 
	summarise(hhlds = survey_total(hhwt)) %>% 
	left_join(count_hhlds, by = c("year", "race2", "sex", "ownershp", "fpl")) %>% 
	mutate(share = round(hhlds / total_hhlds, 3)) %>% 
	select(-contains("total")) %>% 
	select(region, everything())

burden <- bind_rows(rent_burd, own_burd)

state_out$burden_race_gender_tenure_fpl <- burden

#charts
#write individual tables for each disaggregation?


```

## PUMA-level

Cost burden by each of

* race/ethnicity
* gender
* tenure
* percent FPL
* any disability recorded

```{r}

```

## Town-level

Cost burden by tenure

```{r}
#getacs
```

## Disability

Cost burden by tenure for households where at least one household member has a disability. Not sure of the n yet but if I can do this by PUMA, I will.