---
title: "Split out disability analysis from households desiring housing"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r SHH! THIS IS A LIBRARY!}
library(ipumsr)
library(tidyverse)
library(srvyr)
library(tidycensus)
library(hrbrthemes)
library(camiller)
library(scales)
library(kableExtra)
```

```{r gg etc}
set.seed(13)

theme_set(theme_ipsum_rc())

pal <- c("#92c9c5", "#ffe37e", "#ffac8f", "#cb94c4", "#5e63af")

scale_fill_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_fill_manual(values = rev(pal))
  } else {
    scale_fill_manual(values = pal)
  }
}

scale_color_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_color_manual(values = rev(pal))
  } else {
    scale_color_manual(values = pal)
  }
}
```

Tl;dr for code below: I'm making some sociodemographic groups here, including race (white, Black, Latino, Other), some income groups (very low, low, mid low, mid high, high) based on median household income, and cost burden (no burden, cost burden, severe cost burden) using a cost ratio (column called ratio) of monthly housing costs to monthly income.

```{r pums prep}
minc <- get_acs(
  	geography = "county",
	  table = "B19013",
  	state = 09,
	  cache_table = T) %>% 
	arrange(GEOID) %>% 
	mutate(countyfip = seq(from = 1, to = 15, by = 2),
				 name = str_remove(NAME, ", Connecticut")) %>% 
	select(countyfip, name, minc = estimate)

ddi <- read_ipums_ddi("../input_data/usa_00043.xml")

pums <- read_ipums_micro(ddi, verbose = F)  %>% 
	mutate_at(vars(YEAR, PUMA, OWNERSHP, OWNERSHPD, RACE, RACED, HISPAN, HISPAND, DIFFREM, DIFFPHYS, DIFFMOB, DIFFCARE, DIFFEYE, DIFFHEAR), as_factor) %>% 
	mutate_at(vars(PERWT, HHWT), as.numeric) %>% 
	mutate_at(vars(HHINCOME, OWNCOST, RENTGRS, OCC), as.integer) %>% 
	janitor::clean_names() %>% 
	left_join(minc, by = "countyfip") %>% 
	mutate(ratio = hhincome / minc) %>% 
	mutate(
		inc_band = cut(
			ratio,
			breaks = c(-Inf, 0.3, 0.5, .8, 1.2, Inf),
			labels = c("Very low", "Low", "Mid-low", "Mid-high", "High"),
			include.lowest = T, right = T)) %>% 
	mutate(
		inc_band = as.factor(inc_band) %>%
			fct_relevel(., "Very low", "Low", "Mid-low", "Mid-high", "High")) %>% 
	mutate(cb = if_else(ownershp == "Rented", (rentgrs * 12) / hhincome, 99999)) %>% 
	mutate(cb = if_else(ownershp == "Owned or being bought (loan)", (owncost * 12) / hhincome, cb)) %>%
	# if housing cost is 0 and income is 0, no burden
	mutate(cb = if_else((rentgrs == 0 & hhincome == 0), 0, cb)) %>%
	mutate(cb = if_else((owncost == 0 & hhincome == 0), 0, cb)) %>%
	#if income is <=0 and housing cost is >0, burden
	mutate(cb = if_else((rentgrs > 0 & hhincome <= 0), 1, cb)) %>%
	mutate(cb = if_else((owncost > 0 & hhincome <= 0), 1, cb)) %>%
	# some people pay more than 100% income to housing, but I will code these as 1
	mutate(cb = if_else(cb > 1, 1, cb)) %>%
	mutate(
		cost_burden = cut(
			cb,
			breaks = c(-Inf, .3, .5, Inf),
			labels = c("No burden", "Cost-burdened", "Severely cost-burdened"),
			include.lowest = T, right = F)) %>% 
		mutate(race2 = if_else(hispan == "Not Hispanic", as.character(race), "Latino")) %>% 
	mutate(race2 = as.factor(race2) %>% 
				 	fct_recode(Black = "Black/African American/Negro") %>%
				 	fct_other(keep = c("White", "Black", "Latino"), other_level = "Other race") %>%
				 	fct_relevel("White", "Black", "Latino", "Other race"))
```

**John** this is where the change is probably needed. Right now, I have a logical flag that searches across the disability columns and creates a new logical variable that says whether any disability is recorded.

```{r}
#logical flag for any disability
pums$has_disability <- apply(pums, 1, function(x) any(grep("Has|Yes", x)))
```

We want 6 categories to group by. These are recorded in the PUMS data as below:

* DIFFREM = cognitive disability
* DIFFPHYS = physical disability
* DIFFMOB = ambulatory disability
* DIFFCARE = self care/idependent living difficuly
* DIFFEYE = vision disability
* DIFFHEAR = hearing disability

I think the approach I would use is to turn each of those columns into Y/N or T/F rather than "Has (disability)" whatever text label they use.

```{r}
# John
# I am regarding N/A as no disability -- revisit?
pums$diffrem_TF <- ifelse(str_detect(pums$diffrem, "N/A"), NA, 
                          str_detect(pums$diffrem, "Has"))
pums$diffphys_TF <- ifelse(str_detect(pums$diffphys, "N/A"), NA, 
                          str_detect(pums$diffphys, "Has")) 
pums$diffmob_TF <- ifelse(str_detect(pums$diffmob, "N/A"), NA, 
                          str_detect(pums$diffmob, "Has")) 
pums$diffcare_TF <- ifelse(str_detect(pums$diffcare, "N/A"), NA, 
                          str_detect(pums$diffcare, "Yes")) 
pums$diffeye_TF <- ifelse(str_detect(pums$diffeye, "N/A"), NA, 
                          str_detect(pums$diffeye, "Yes")) 
pums$diffhear_TF <- ifelse(str_detect(pums$diffhear, "N/A"), NA, 
                          str_detect(pums$diffhear, "Yes")) 
pums$disability_sum <- pums %>% 
  select(diffrem_TF:diffhear_TF) %>% 
  apply(1, sum, na.rm=TRUE)
```


The chunk below pulls the data together by creating a LUT of households with any occupant with a disability, so we only count the household once. You'll need to set it up so we count each household only once per disability type. In other words, we can have a household counted more than once if they have an occupant with more than one disability type, or if they have multiple occupants with different disabilities.

```{r setup data}
des <- pums %>%
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>%
	as_survey_design(., ids = 1, wt = hhwt)

ct_hhlds <- des %>%
	select(hhwt, statefip, inc_band) %>% 
	group_by(statefip, inc_band) %>% 
	summarise(value = survey_total(hhwt)) %>% 
	mutate(name = "Connecticut", level = "1_state") %>% 
	select(-statefip)

hh_w_disability <- pums %>%
	filter(has_disability == T) %>% 
	select(cbserial) %>% #cbserial is the hh code
	unique()

ct_inc_band_disability <- des %>% 
	mutate(disability = if_else(cbserial %in% hh_w_disability$cbserial, T, F)) %>% 
	select(hhwt, inc_band, disability) %>% 
	mutate(name = "Connecticut") %>% 
	group_by(name, inc_band, disability) %>% 
	summarise(value = survey_total(hhwt)) %>% 
	ungroup() %>% 
	group_by(name, inc_band)

ct_total_disability <- ct_inc_band_disability %>% 
	select(-value_se) %>% 
  ungroup() %>% 
	mutate(inc_band = "Total") %>% 
	group_by(name, inc_band, disability) %>% 
	summarise(value = sum(value))

ct_inc_band_disability <- ct_inc_band_disability %>% 
	bind_rows(ct_total_disability)
```


```{r setup data}
# John
des <- pums %>%
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>%
	as_survey_design(., ids = 1, wt = hhwt)

ct_hhlds_no_total <- des %>%
	select(hhwt, statefip, inc_band) %>% 
	group_by(statefip, inc_band) %>% 
	summarise(value = survey_total(hhwt)) %>% 
	mutate(name = "Connecticut", level = "1_state") %>% 
	select(-statefip)

ct_hhlds_total <- ct_hhlds_no_total %>% 
    select(-value_se) %>% 
    ungroup() %>% 
    mutate(inc_band = "Total") %>%
    group_by(name, inc_band) %>% 
    summarise(value = sum(value))

ct_hhlds <- ct_hhlds_no_total %>% 
  bind_rows(ct_hhlds_total)

get_counts <- function(col) {
  colname <- as.name(col)
  inc_band_disability <- des %>% 
    select(hhwt, inc_band, !!colname) %>% 
  	mutate(name = "Connecticut") %>% 
    mutate(disability_type = col %>% str_remove("_TF")) %>% 
  	group_by(name, inc_band, disability_type, !!colname) %>% 
  	summarise(value = survey_total(hhwt)) %>% 
  	ungroup() %>% 
    select(name, disability_type, inc_band, disability=!!colname, value, value_se) %>% 
  	group_by(name, inc_band, disability_type)
  total_disability <- inc_band_disability %>% 
    select(-value_se) %>% 
    ungroup() %>% 
    mutate(inc_band = "Total") %>%
    group_by(name, disability_type, inc_band, disability) %>% 
    summarise(value = sum(value))
  ct_inc_band_disability <- inc_band_disability %>% 
    bind_rows(total_disability)
  return(ct_inc_band_disability)
}

tf_vars <- pums %>% 
  select(ends_with("TF")) %>% 
  names()

ct_inc_band_disability_type <- tf_vars %>% 
  map_dfr(get_counts)
```


My original tables and charts are below, with comments in **bold** regarding new vizzes.

Quick tabulations of households with an occupant with a disability

**We want this table to show households by income band by disability type**
```{r}
kable(ct_inc_band_disability)
```
```{r}
# John
tf_vars %>% 
  lapply(get_counts) %>% 
  lapply(kable) # separate tables for each disability type
```



Taking a slightly different look than the table above, the plot below shows the share of households in each income band that have an occupant with a disability (so the numerator is households with an occupant with a disability in a given income band, and the denominator is all households in that income band).

While the High band has the most (by count) households with an occupant with a disability, it has the smallest share.

**I think we might want a plot faceted by disability type.**

```{r}
ct_inc_band_disability %>%
	bind_rows(ct_hhlds) %>% 
	ungroup() %>% 
	select(-level) %>% 
	filter(inc_band != "Total") %>% 
	mutate(inc_band = as.factor(inc_band) %>% 
				 	fct_relevel(., "Very low", "Low", "Mid-low", "Mid-high", "High") %>% 
				 	fct_rev()) %>%
	mutate(disability = as.character(disability)) %>% 
	mutate(disability = if_else(is.na(disability), "Total", disability)) %>% 
	group_by(name, inc_band) %>% 
	calc_shares(group = disability, denom = "Total", value = value, moe = value_se) %>% 
	filter(disability == "TRUE", inc_band != "Total") %>% 
	ggplot(aes(share, inc_band, group = inc_band)) +
	geom_col(aes(fill = inc_band), position = position_dodge(1)) +
	geom_text(aes(label = percent(share, accuracy = 1)), position = position_dodge(1), family = "Roboto Condensed", size = 4.5, hjust = 1.1, vjust = .5) +
	scale_x_continuous(expand = expansion(mult = c(0, 0))) +
#	scale_fill_custom(rev = T) + 
	theme(plot.title.position = "plot",
				axis.text.x = element_blank(), 
				axis.text.y = element_text(colour = "black"),
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				legend.position = "none") +
	labs(title = str_wrap("Share of households with a person with a disability in each income band", 60),
			 subtitle = "Connecticut",
			 x = "", y = "")
```

```{r}
# John
ct_inc_band_disability_type %>%
	bind_rows(ct_hhlds %>% mutate(disability_type = "diffrem")) %>% # messy...
  bind_rows(ct_hhlds %>% mutate(disability_type = "diffphys")) %>% 
  bind_rows(ct_hhlds %>% mutate(disability_type = "diffmob")) %>% 
  bind_rows(ct_hhlds %>% mutate(disability_type = "diffcare")) %>% 
  bind_rows(ct_hhlds %>% mutate(disability_type = "diffeye")) %>% 
  bind_rows(ct_hhlds %>% mutate(disability_type = "diffhear")) %>% 
	ungroup() %>%
	select(-level) %>% 
	mutate(inc_band = as.factor(inc_band) %>% 
				 	fct_relevel(., "Very low", "Low", "Mid-low", "Mid-high", "High", "Total") %>% 
				 	fct_rev()) %>%
	mutate(disability = as.character(disability)) %>% 
	mutate(disability = if_else(is.na(disability), "Total", disability)) %>% 
	group_by(name, disability_type, inc_band) %>% 
	calc_shares(group = disability, denom = "Total", value = value, moe = value_se) %>% 
	filter(disability == "TRUE", inc_band != "Total") %>% 
	ggplot(aes(share, inc_band, group = disability_type)) +
	geom_col(position = position_dodge(1)) +
  facet_wrap(~ disability_type) + 
	geom_text(aes(label = percent(share, accuracy = 1)), position = position_dodge(1), size = 4.5, hjust = 1.1, vjust = .5) +
	scale_x_continuous(expand = expansion(mult = c(0, 0))) +
#	scale_fill_custom(rev = T) + # temp. commented bcus John was having ggplot errors
	theme(plot.title.position = "plot",
				axis.text.x = element_blank(), 
				axis.text.y = element_text(colour = "black"),
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				legend.position = "none") +
	labs(title = str_wrap("Share of households with a person with a disability in each income band", 60),
			 subtitle = "Connecticut",
			 x = "", y = "")
```



Finally, what are the average cost burden rates for households with an occupant with a disability?

```{r}
ct_cost_ratio_inc_band_disability <- pums %>% 
	mutate(disability = if_else(cbserial %in% hh_w_disability$cbserial, T, F)) %>% 
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>% 
	select(hhwt, inc_band, disability, cb) %>% 
	mutate(mult = cb * hhwt) %>% 
	group_by(inc_band, disability) %>% 
	summarise(avg_cost_ratio = sum(mult) / sum(hhwt)) %>% 
	ungroup() %>% 
	mutate(name = "Connecticut")
```

Read this chart as: "Households with an occupant with a disability in the Very Low cost band spend, on average, 66% of household income on housing costs." 

So the average cost ratio for households with an occupant with a disability is slightly lower than households without a disabled occupant, but the general trends in cost ratio across the cost bands still hold.

**I think this chart is fine as it is, and I just included it so this notebook has all the disability tabuations in it.**

```{r}
ct_cost_ratio_inc_band_disability %>% 
	ggplot(aes(inc_band, avg_cost_ratio, group = disability)) +
	geom_col(aes(fill = disability), width = .8, position = position_dodge(.85)) +
	geom_text(aes(label = percent(avg_cost_ratio, accuracy = 1)), position = position_dodge(.85), family = "Roboto Condensed", hjust = .5, vjust = 1.2) +
	scale_fill_manual(values = c(pal[1], pal [4])) +
	guides(fill = guide_legend(title = "Household has occupant with disability")) +
	labs(title = "Average cost ratio for households in each income band",
			 subtitle = "Connecticut",
			 x = "", y = "",
			 caption = "Cost ratio is the amount of household income spent on housing costs.") +
	theme(plot.title.position = "plot", 
				legend.position = "bottom",
				panel.grid.minor = element_blank(),
				panel.grid.major = element_blank(),
				axis.text.y = element_blank(),
				axis.text.x = element_text(colour = "black"))
```