---
title: "Housing units"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(cwi)
library(camiller)
library(ipumsr)
library(srvyr)
```

Change in total units by tenure (occ + vacant?), units in structure (SF, 2-4 units, 5+ units), number bedrooms. Presumably PUMS because ACS is households not units. DECD data has vacancy among total units for 2010. Do we want 2000, 2018?

## Setup PUMS file
```{r pums}
names <- tibble(countyfip = seq(from = 1, to = 15, by = 2),
								name = c("Fairfield County", "Hartford County", "Litchfield County", "Middlesex County", "New Haven County", "New London County", "Tolland County", "Windham County"))

ddi <- read_ipums_ddi("../input_data/usa_00047.xml")

pums <- read_ipums_micro(ddi, verbose = F) %>% 
	filter(RECTYPE == "H") %>% 
	mutate_at(vars(YEAR, OWNERSHP, VACANCY, UNITSSTR, BEDROOMS), as_factor) %>% 
	mutate_at(vars(HHWT, COUNTYFIP), as.numeric) %>% 
	clean_names() %>% 
	left_join(names, by = "countyfip") %>% 
	select(year, name, ownershp, vacancy, unitsstr, bedrooms, hhwt) %>% 
	mutate(ownershp2 = fct_recode(ownershp,
																`Owner` = "Owned or being bought (loan)",
																`Renter` = "Rented"),
				 ownershp2 = as.character(ownershp2),
				 ownershp2 = if_else(vacancy == "For rent or sale", "Renter", ownershp2),
				 ownershp2 = if_else(vacancy == "For sale only", "Owner", ownershp2),
				 ownershp2 = as.factor(ownershp2)) %>% 
	mutate(type = as.factor(unitsstr) %>% 
				 	fct_collapse(., `Single family` = c("1-family house, detached", "1-family house, attached", "Mobile home or trailer", "Boat, tent, van, other"),
				 							 Multifamily = c("2-family building", "3-4 family building", "5-9 family building", "10-19 family building", "20-49 family building", "50+ family building"))) %>% 
	mutate(bdr = as.factor(bedrooms) %>% 
				 	fct_collapse(., `Studio or 1 bedroom` = c("No bedrooms", "1"),
				 							 `2 bedrooms` = "2",
				 							 `3 bedrooms` = "3",
				 							 `4 or more bedrooms` = c("4 (1970-2000, 2000-2007 ACS/PRCS)", "5+ (1970-2000, 2000-2007 ACS/PRCS)", "6", "7", "8")))
```

# By tenure

```{r}
tenure <- pums %>% 
	filter(ownershp2 != "N/A") %>% 
	select(year, name, ownershp2, hhwt) %>% 
	group_by(year, name, ownershp2) %>% 
	summarise(units = sum(hhwt)) %>% 
	ungroup()

write_csv(tenure, path = "../output_data/units_by_tenure_2000_2018.csv")
```

```{r, fig.height=8, fig.width=8}
tenure %>% 
	filter(year != 2010) %>% 
	pivot_wider(id_cols = c(name, ownershp2), names_from = year, names_prefix = "x", values_from = units) %>% 
	mutate(diff = x2018 - x2000) %>% 
	mutate(name = as.factor(name) %>% fct_rev(),
				 ownershp2 = fct_rev(ownershp2)) %>% 
	ggplot(aes(diff, name, group = ownershp2)) +
	geom_col(aes(fill = ownershp2), width = .8, position = position_dodge(.85)) +
	geom_text(aes(label = scales::comma(diff, accuracy = 1)), hjust = 1.04,
						position = position_dodge(.85), family = "Roboto Condensed") +
	scale_x_continuous(labels = scales::comma_format(),
										 expand = expansion(mult = c(.1, .1))) +
	guides(fill = guide_legend(title = "", reverse = T)) +
	hrbrthemes::theme_ipsum_rc() +
	labs(title = "Change in total units by tenure, 2000–2018",
			 x = "", y = "",
			 caption = "Includes vacant units.") +
	theme(plot.title.position = "plot",
				legend.position = "right", 
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_blank(),
				axis.text.y = element_text(colour = "black"))
```

## By building type

Small multifamily (2-4 units) are so small it messes up the scale of the bars, so I'm lumping all MF together

```{r}
bldg_type <- pums %>% 
	filter(type != "N/A") %>% 
	select(year, name, type, hhwt) %>% 
	group_by(year, name, type) %>% 
	summarise(units = sum(hhwt)) %>% 
	ungroup()

write_csv(bldg_type, path = "../output_data/units_by_bldg_type_2000_2018.csv")
```

```{r, fig.height=8, fig.width=8}
bldg_type %>% 
	filter(year != 2010) %>% 
	pivot_wider(id_cols = c(name, type), names_from = year, names_prefix = "x", values_from = units) %>% 
	mutate(diff = x2018 - x2000) %>% 
	mutate(name = as.factor(name) %>% fct_rev(),
				 type = fct_rev(type)) %>% 
	ggplot(aes(diff, name, group = type)) +
	geom_col(aes(fill = type), width = .8, position = position_dodge(.85)) +
	geom_text(aes(label = scales::comma(diff, accuracy = 1)), hjust = 1.04,
						position = position_dodge(.85), family = "Roboto Condensed") +
	scale_x_continuous(labels = scales::comma_format(),
										 expand = expansion(mult = c(.05, .1))) +
	guides(fill = guide_legend(title = "", reverse = T)) +
	hrbrthemes::theme_ipsum_rc() +
	labs(title = "Change in total units by building type, 2000–2018",
			 x = "", y = "",
			 caption = "Includes vacant units.") +
	theme(plot.title.position = "plot",
				legend.position = "right", 
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_blank(),
				axis.text.y = element_text(colour = "black"))
```

## By number of bedrooms

0-1, 2, 3, 4+

```{r}
bedrooms <- pums %>% 
	filter(bdr != "N/A") %>% 
	select(year, name, bdr, hhwt) %>% 
	group_by(year, name, bdr) %>% 
	summarise(units = sum(hhwt)) %>% 
	ungroup()

write_csv(bedrooms, path = "../output_data/units_by_bedrooms_2000_2018.csv")
```

```{r, fig.height=10, fig.width=10}
bedrooms %>% 
	filter(year != 2010) %>% 
	pivot_wider(id_cols = c(name, bdr), names_from = year, names_prefix = "x", values_from = units) %>% 
	mutate(diff = x2018 - x2000) %>% 
	mutate(name = as.factor(name) %>% fct_rev(),
				 bdr = fct_rev(bdr)) %>% 
	ggplot(aes(diff, name, group = bdr)) +
	geom_col(aes(fill = bdr), width = .8, position = position_dodge(.85)) +
	geom_text(aes(label = scales::comma(diff, accuracy = 1)), hjust = 1.04,
						position = position_dodge(.85), family = "Roboto Condensed") +
	scale_x_continuous(labels = scales::comma_format(),
										 expand = expansion(mult = c(.05, .1))) +
	guides(fill = guide_legend(title = "", reverse = T)) +
	hrbrthemes::theme_ipsum_rc() +
	labs(title = "Change in total units by building type, 2000–2018",
			 x = "", y = "",
			 caption = "Includes vacant units.") +
	theme(plot.title.position = "plot",
				legend.position = "bottom", 
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_blank(),
				axis.text.y = element_text(colour = "black"))
```