---
title: "Median income by county and tenure"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(ipumsr)
library(srvyr)
```

```{r gg etc}
theme_set(hrbrthemes::theme_ipsum_rc(base_family = "Lato Regular"))

#urban colors
pal <- c("#1696d2", "#fdbf11", "#d2d2d2", "#ec008b", "#55b748")

scale_fill_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_fill_manual(values = rev(pal))
  } else {
    scale_fill_manual(values = pal)
  }
}

scale_color_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_color_manual(values = rev(pal))
  } else {
    scale_color_manual(values = pal)
  }
}
```

Two ways: 

Median income by county. Simple enough.

Median income by tenure. I thought about also doing this by race, but tenure makes more sense given the rest of the analysis in the housing report. We can draw on other DataHaven analyses on median income by race if we want to mention it (white median income is nearly 50% higher than Black and Latino median income—pull data from HER notebooks).

## Setup PUMS file

```{r}
names <- tibble(countyfip = seq(from = 1, to = 15, by = 2),
								name = c("Fairfield County", "Hartford County", "Litchfield County", "Middlesex County", "New Haven County", "New London County", "Tolland County", "Windham County"))

infl_values <- blscrapeR::inflation_adjust(base_year = 2018) %>% 
	filter(year %in% c(2000, 2010, 2018)) %>% 
	arrange(desc(year)) %>% 
	mutate(factor = 1 + (abs(pct_increase) / 100)) %>% 
	select(year, factor)

ddi <- read_ipums_ddi("../input_data/usa_00049.xml")

pums <- read_ipums_micro(ddi, verbose = F) %>% 
	mutate_at(vars(YEAR, OWNERSHP, OWNERSHPD), as_factor) %>% 
	mutate_at(vars(HHINCOME, TRANTIME), as.numeric) %>% 
	janitor::clean_names() %>% 
	left_join(names, by = "countyfip") %>% 
	left_join(infl_values, by = "year") %>% 
	mutate(adj_income = hhincome * factor) %>% 
	mutate(ownershp2 = if_else(ownershp == "Rented", "Renter", "Owner")) %>% 
	select(year, pernum, hhwt, name, ownershp, ownershp2, hhincome, adj_income)
```

## By county and state, no disaggregation.

The Bureau is migrating its API endpoints so all censusapi and tidycensus are both struggling with the decennial call. I'll use PUMS data for this and round it to the nearest thousand to accommodate any difference in median calculation between deci/ACS and microdata.

```{r}
des <- pums %>%
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>% 
	as_survey_design(., ids = 1, wt = hhwt)

county_minc <- des %>%
	select(name, year, adj_income) %>% 
	group_by(name, year) %>% 
	summarise(minc = survey_median(adj_income)) %>% 
	mutate(minc = round(minc, 0))

state_minc <- des %>% 
	mutate(name = "Connecticut") %>% 
	select(name, year, adj_income) %>% 
	group_by(name, year) %>% 
	summarise(minc = survey_median(adj_income)) %>% 
	mutate(minc = round(minc, 0))

minc <- bind_rows(state_minc, county_minc)

write_csv(minc, "../output_data/minc_by_county_2000_2018.csv")
```

```{r fig.height=10, fig.width=10}
minc %>% 
	ggplot(aes(year, minc)) +
	geom_vline(aes(xintercept = year), size = .5, color = "grey70") +
	geom_point(size = 4, color = pal[1]) +
	geom_line(size = 1, color = pal[1]) +
	geom_text(aes(label = scales::dollar(minc, accuracy = 1e3), vjust = 1.6, family = "Lato Regular", nudge_y = -500)) +
	#scale_y_continuous(expand = expansion(mult = c(.15, .15))) +
	scale_y_continuous(limits = c(50000, 105000)) +
	facet_wrap(facets = "name", scales = "free_y") +
	hrbrthemes::theme_ipsum_rc() +
	labs(title = "Median income, 2000–2018",
			 x = "", y = "") +
	theme(plot.title.position = "plot",
				axis.text.y = element_blank(),
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_text(colour = "black"),
				strip.text.x = element_text(hjust = .5),
				legend.position = "none")
```

```{r}
minc %>% 
	select(name, year, `median income` = minc) %>% 
	kableExtra::kable()
```

## By tenure

```{r}
des2 <- pums %>%
	filter(pernum == "1", hhincome != 9999999, ownershp != "N/A") %>% 
	as_survey_design(., ids = 1, wt = hhwt)

county_minc2 <- des2 %>%
	select(name, year, ownershp2, adj_income) %>% 
	group_by(name, year, ownershp2) %>% 
	summarise(minc = survey_median(adj_income)) %>% 
	mutate(minc = round(minc, 0))

state_minc2 <- des2 %>% 
	mutate(name = "Connecticut") %>% 
	select(name, year, ownershp2, adj_income) %>% 
	group_by(name, year, ownershp2) %>% 
	summarise(minc = survey_median(adj_income)) %>% 
	mutate(minc = round(minc, 0))

minc_tenure <- bind_rows(state_minc2, county_minc2)

write_csv(minc_tenure, "../output_data/minc_by_tenure_2000_2018.csv")
```

```{r fig.height=10, fig.width=10}
minc_tenure %>% 
	ggplot(aes(year, minc, group = ownershp2)) +
	geom_vline(aes(xintercept = year), size = .5, color = "grey70") +
	geom_point(aes(color = ownershp2), size = 4) +
	geom_line(aes(color = ownershp2), size = 1) +
	geom_text(aes(label = scales::dollar(minc, accuracy = 1e3), vjust = 1.6, family = "Lato Regular")) +
	scale_y_continuous(expand = expansion(mult = c(.35, .25))) +
	#scale_y_continuous(limits = c(0, 130000)) +
	facet_wrap(facets = "name", scales = "free_y") +
	hrbrthemes::theme_ipsum_rc() +
  scale_color_manual(values = c(pal[1:2])) +
	guides(color = guide_legend(title = "Tenure", override.aes = list(linetype = 0))) +
	labs(title = "Median income by tenure and area, 2000–2018",
			 x = "", y = "") +
	theme(plot.title.position = "plot",
				axis.text.y = element_blank(),
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_text(colour = "black"),
				strip.text.x = element_text(hjust = .5),
				legend.position = "bottom")
```

```{r}
minc_tenure %>% 
	select(name, year, `median income` = minc) %>% 
	kableExtra::kable()
```