---
title: "Pop by race"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(cwi)
library(camiller)
```

```{r gg etc}
theme_set(hrbrthemes::theme_ipsum_rc(base_family = "Lato Regular"))

#urban colors
pal <- c("#1696d2", "#fdbf11", "#d2d2d2", "#ec008b", "#55b748")

scale_fill_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_fill_manual(values = rev(pal))
  } else {
    scale_fill_manual(values = pal)
  }
}

scale_color_custom <- function(palette = pal, rev = F) {
  if (rev) {
    scale_color_manual(values = rev(pal))
  } else {
    scale_color_manual(values = pal)
  }
}
```

Collecting and lightly cleaning basic population data for multiple
geographies and each year available starting in 2000 through latest
available.

Using four major race/ethnicity groups and all others grouped together, for five groups total - state and county intercensal (2001-2009), decennial (2000 and 2010), and ACS 2011-2018; town ACS 2011-2018 and decinnial 2000 and 2010

2000-2010 intercensal data comes in spreadsheets from
<https://www.census.gov/data/datasets/time-series/demo/popest/intercensal-2000-2010-counties.html>

See data dictionary in companion pdf in input_data.

## Fetch

```{r}
intercensal <- read_csv("../input_data/co-est00int-alldata-09.csv") %>%
	clean_names()

acs_years <- list("2011" = 2011, "2012" = 2012, "2013" = 2013, "2014" = 2014, "2015" = 2015, "2016" = 2016, "2017" = 2017, "2018" = 2018)

race <- acs_years %>% map(~multi_geo_acs(table = "B03002", year = ., new_england = F, verbose = F))
race_bind <- Reduce(rbind, race) %>% label_acs()

# Decennial race/ethnicity tables are different, need to do two calls :weary:
race_fetch00 <- multi_geo_decennial(table = "P004", year = 2000, verbose = F) %>% 
	label_decennial(year = 2000)

race_fetch10 <- multi_geo_decennial(table = "P005", year = 2010, verbose = F) %>% 
	label_decennial()
```

## Clean

Intercensals have estimates that are a little off. Other race is too small while Asian and Latino are a bit too high compared with decennial census figures. Use at your own peril.

```{r}
deci_race10 <- race_fetch10 %>% 
	clean_names() %>% 
	group_by(level, geoid, name, year) %>%
  add_grps(list(total_pop = 1, latino = 10, white = 3, black = 4, asian = 6, other_race = c(5, 7:9)), group = label, value = value) %>%
  calc_shares(group = label, value = value) %>%
	mutate(moe = 0, sharemoe = NA) %>% 
  rename(var = label, estimate = value)

deci_race00 <- race_fetch00 %>% 
	clean_names() %>% 
	group_by(level, geoid, name, year) %>%
  add_grps(list(total_pop = 1, latino = 2, white = 5, black = 6, asian = 8, other_race = c(7, 9:11)), group = label, value = value) %>%
  calc_shares(group = label, value = value) %>%
	mutate(moe = 0, sharemoe = NA) %>% 
  rename(var = label, estimate = value)

deci_race <- bind_rows(deci_race00, deci_race10)

acs_race <- race_bind %>%
	clean_names() %>% 
	group_by(level, geoid, name, year, county) %>% 
  add_grps(list(total_pop = 1, latino = 12, white = 3, black = 4, asian = 6, other_race = c(5, 7:9)), group = label, moe = moe) %>%
  calc_shares(group = label, moe = moe) %>%
  rename(var = label)

period_lut <- tibble(
	estimate_date = c(
		"remove_april_2000",
		"remove_july_2000",
		seq(2001, 2009),
		"remove_april_2010",
		"remove_july_2010"),
	year = seq(1:13))

int_race_county <- intercensal %>% 
	filter(agegrp == 99) %>% 
	mutate(geoid = paste(state, county, sep = ""),
				 latino = h_male + h_female,
				 white = nhwa_male + nhwa_female,
				 black = nhba_male + nhba_female,
				 asian = nhaa_male + nhaa_female,
				 other_race = (nh_male + nh_female) - (white + black + asian)) %>% 
	select(geoid, ctyname, year, tot_pop, white, black, latino, asian, other_race) %>% 
	left_join(., period_lut, by = "year") %>% 
	filter(!grepl("remove", estimate_date)) %>% 
	select(-year) %>% 
	rename(total_pop = tot_pop, year = estimate_date, name = ctyname) %>% 
	mutate(year = as.numeric(year),
				 level = "2_counties",
				 level = as.factor(level)) %>% 
	select(year, geoid, name, level, everything()) %>% 
	pivot_longer(cols = 5:10, names_to = "var", values_to = "estimate") %>% 
	group_by(year, geoid, name, level) %>% 
	calc_shares(group = var, value = estimate, denom = "total_pop") %>% 
  ungroup()

int_race_ct <- int_race_county %>%
	select(-name, -level, -geoid, -share) %>% 
	mutate(name = "Connecticut", level = "1_state", geoid = "09",
				 level = as.factor(level)) %>% 
	group_by(geoid, name, level, year, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	group_by(geoid, name, level, year) %>% 
	calc_shares(group = var, value = estimate) %>% 
	ungroup()

pop_by_race_out <- bind_rows(deci_race, acs_race, int_race_county, int_race_ct) %>% 
	select(level, geoid, name, year, county, var, estimate, moe, share, sharemoe) %>% 
	arrange(level, geoid, year)

write_csv(pop_by_race_out, "../output_data/pop_by_race_2000_2018.csv")
```

## Calculate change
```{r}
race_change <- pop_by_race_out %>%
	select(-share, -moe, -sharemoe) %>%
	rename(race = var) %>% 
	group_by(level, geoid, county, race) %>%
	arrange(name, year, race) %>%
	mutate(diff = estimate - lag(estimate, default = first(estimate))) %>%
	arrange(level, geoid, year, race) %>%
	mutate(measure = "pop_change_from_prev_data_year") %>%
	select(-estimate) %>%
	rename(estimate = diff) %>% 
	select(year, level, geoid, name, county, race, measure, estimate)
	
race_change %>% 
	write_csv("../output_data/pop_by_race_change_2000_2018.csv")
```

```{r fig.height=8, fig.width=8}
labels <- tibble(var = c("black", "white", "latino", "other_race", "total_pop", "asian"),
								 label = c("Black, non-Latino", "White, non-Latino", "Latino", "Other race, non-Latino", "Total population", "Asian, non-Latino"))

pop_by_race_out <- read_csv("../output_data/pop_by_race_2000_2018.csv")

pop_by_race_out %>% 
	ungroup() %>% 
	filter(level != "3_towns", year %in% c(2000, 2010, 2018)) %>% 
	mutate(var = fct_collapse(var, `other_race` = c("asian", "other_race"))) %>% 
	select(year, name, var, estimate) %>% 
	group_by(year, name, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	group_by(year, name) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate) %>% 
	ungroup() %>% 
	mutate(year = as.factor(year)) %>% 
	filter(var != "total_pop") %>% 
	left_join(labels, by = "var") %>% 
	mutate(label = as.factor(label) %>% 
				 	fct_relevel(., "White, non-Latino", "Black, non-Latino", "Latino", "Other race, non-Latino")) %>% 
	ggplot(aes(year, share, group = label)) +
	geom_point(aes(color = label), size = 3, alpha = .8) +
	geom_line(aes(color = label), size = .75, alpha = .8) +
	scale_y_continuous(labels = scales::percent_format()) +
	scale_x_discrete(expand = expansion(mult = c(0.05,0.05))) +
	facet_wrap(facets = "name") +
#	hrbrthemes::theme_ipsum_rc() +
  scale_color_manual(values = c(pal[5], pal[4], pal[2], pal[1])) +
	guides(color = guide_legend(title = "", override.aes = list(linetype = 0))) +
	labs(title = "Population by race/ethnicity, 2000–2018",
			 x = "", y = "") +
	theme(plot.title.position = "plot",
				axis.text.y = element_text(colour = "black"),
				strip.text.x = element_text(hjust = .5),
				panel.grid.minor = element_blank(),
				axis.text.x = element_text(colour = "black"),
				legend.position = "bottom")
```

```{r}
pop_by_race_out %>% 
	ungroup() %>% 
	filter(level != "3_towns", year %in% c(2000, 2010, 2018)) %>% 
	mutate(var = fct_collapse(var, `other_race` = c("asian", "other_race"))) %>% 
	select(year, name, var, estimate) %>% 
	group_by(year, name, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	left_join(labels, by = "var") %>% 
	select(-var) %>% 
	pivot_wider(id_cols = c("name", "label"), names_from = year, values_from = estimate) %>% 
	mutate(pct_chg = (`2018` - `2000`) / `2000`,
				 lbl_chg = scales::percent(pct_chg, accuracy = 1),
				 `2000` = scales::comma(`2000`, accuracy = 1),
				 `2018` = scales::comma(`2018`, accuracy = 1)) %>% 
	select(Name = name, `Race/ethnic group` = label, `Estimate, 2000` = `2000`, `Estimate, 2018` = `2018`, `Pct. chg. 2000–2018` = lbl_chg) %>% 
	write_csv(file = "../output_data/pop_race_change_percent_2000_2018.csv") %>% 
	kableExtra::kable()
```

```{r fig.height=8, fig.width=8}
pop_by_race_out %>% 
	ungroup() %>% 
	filter(level != "3_towns", year %in% c(2000, 2010, 2018)) %>% 
	mutate(var = fct_collapse(var, `other_race` = c("asian", "other_race"))) %>% 
	select(year, name, var, estimate) %>% 
	group_by(year, name, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	left_join(labels, by = "var") %>% 
	select(-var) %>% 
	mutate(label = as.factor(label) %>% 
				 	fct_relevel(., "Total population", "White, non-Latino", "Black, non-Latino", "Latino", "Other race, non-Latino") %>% 
				 	fct_rev()) %>% 
	pivot_wider(id_cols = c("name", "label"), names_from = year, values_from = estimate) %>% 
	mutate(pct_chg = (`2018` - `2000`) / `2000`) %>% 
	ggplot(aes(pct_chg, label)) +
	geom_vline(xintercept = 0, size = .25, alpha = .8) +
	geom_col(aes(fill = label), width = .75, position = position_dodge(.85)) +
	geom_text(aes(label = scales::percent(pct_chg, accuracy = 1)), position = position_dodge(.85), hjust = 1, family = "Lato Regular", size = 4) +
	scale_x_continuous(limits = c(-.25, 2),
										 labels = scales::percent_format(),
										 expand = expansion(mult = c(.15, .1))) +
	facet_wrap(facets = "name") +
#	hrbrthemes::theme_ipsum_rc() +
  scale_fill_manual(values = rep(pal[1], 5)) +
	guides(fill = guide_legend(title = "")) +
	labs(title = "Population change by race/ethnicity, 2000–2018",
			 x = "", y = "") +
	theme(plot.title.position = "plot",
				axis.text.y = element_text(colour = "black"),
				strip.text.x = element_text(hjust = .5),
				panel.grid.minor = element_blank(),
				panel.grid.major = element_blank(),
				axis.text.x = element_blank(),
				legend.position = "none")
```

```{r fig.height=8, fig.width=8}
pop_by_race_out %>% 
	ungroup() %>% 
	filter(year == 2018, level != "3_towns", var != "total_pop") %>% 
	left_join(labels, by = "var") %>% 
	mutate(label = as.factor(label) %>% 
				 	fct_relevel(., "White, non-Latino", "Black, non-Latino", "Latino", "Asian, non-Latino", "Other race, non-Latino") %>% 
				 	fct_rev()) %>% 
	ggplot(aes(estimate, label, group = name)) +
	geom_col(aes(fill = label), width = .75, position = position_dodge(.85)) +
	geom_text(aes(label = scales::comma(estimate, accuracy = 1)), position = position_dodge(.85), hjust = "inward", family = "Lato Regular", size = 4) +
	scale_x_continuous(labels = scales::comma_format(),
										 expand = expansion(mult = c(.15, .1))) +
	facet_wrap(facets = "name", scales = "free_x") +
#	hrbrthemes::theme_ipsum_rc(base_family = "Lato Regular") +
  scale_fill_manual(values = rep(pal[1], 5)) +
	guides(fill = guide_legend(title = "")) +
	labs(title = "Population by race/ethnicity, 2018",
			 x = "", y = "",
			 caption = str_wrap("'Other Race, non-Latino' includes American Indian/Alaska Native, Native Hawaiian/Pacific Islander, people indicating 'Some Other Race,' and people of two or more races who are non-Latino.", 120)) +
	theme(plot.title.position = "plot",
				axis.text.y = element_text(colour = "black"),
				strip.text.x = element_text(hjust = .5),
				panel.grid.minor = element_blank(),
				panel.grid.major = element_blank(),
				axis.text.x = element_blank(),
				legend.position = "none")
```