---
title: "Pop by age"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(cwi)
library(camiller)
```

Collecting and lightly cleaning basic population data for multiple
geographies and each year available starting in 2000 through latest
available.

Using 5-year bands - state, county intercensal and ACS 2000-2018; town ACS 2011-2018 and deci 2010

2000-2010 intercensal data comes in spreadsheets from
<https://www.census.gov/data/datasets/time-series/demo/popest/intercensal-2000-2010-counties.html>

See data dictionary in companion pdf in input_data.

## Fetch

```{r}
intercensal <- read_csv("../input_data/co-est00int-alldata-09.csv") %>%
	clean_names()

acs_years <- list("2011" = 2011, "2012" = 2012, "2013" = 2013, "2014" = 2014, "2015" = 2015, "2016" = 2016, "2017" = 2017, "2018" = 2018)

#b01001 - sex by age
sex_by_age <- acs_years %>% map(~multi_geo_acs(table = "B01001", year = ., new_england = F))
sex_by_age_bind <- Reduce(rbind, sex_by_age) %>% label_acs()

deci_years <- list("2000" = 2000, "2010" = 2010)

deci_pops <- deci_years %>% map(~multi_geo_decennial(table = "P012", year = .))

deci_pops_bind <- Reduce(rbind, deci_pops) %>% label_decennial() 
```

I want to add 1990 to this to match up with the total pop charts elsewhere, but it's a different table in the 1990 Census so needs its own call.

```{r}
pop90_fetch <- multi_geo_decennial(table = "P011", year = 1990) %>% 
	label_decennial(year = 1990) %>% 
	rename(age = label) %>% 
	mutate(age = str_to_lower(age),
				 age = str_remove(age, "persons age ")) %>% 
	mutate(age = as.factor(age) %>% 
				 	fct_collapse(.,
				 							 under_5_years = c("< 1 yr", "1-2 yrs", "3-4 yrs"),
				 							 `5_to_9_years` = c("5 yrs", "6 yrs", "7-9 yrs"),
				 							 `10_to_14_years` = c("10-11 yrs", "12-13 yrs", "14 yrs"),
				 							 `15_to_19_years` = c("15 yrs", "16 yrs", "17 yrs", "18 yrs", "19 yrs"),
				 							 `20_to_24_years` = c("19 yrs", "20 yrs", "21 yrs", "22-24 yrs"),
				 							 `60_to_64_years` = c("60-61 yrs", "62-64 yrs"),
				 							 `85_years_and_over` = c("85+ yrs")),
				 age = str_replace(age, "-", "_to_"),
				 age = str_replace(age, " yrs", "_years"),
				 age = fct_relevel(age, "under_5_years", "5_to_9_years")) %>% 
	clean_names() %>% 
	select(-variable) %>% 
	group_by(year, level, geoid, name, state, county, age) %>% 
	summarise(value = sum(value))

total90 <- pop90_fetch %>% 
	ungroup() %>% 
	select(-age) %>% 
	group_by(year, level, geoid, name, state, county) %>% 
	summarise(value = sum(value)) %>% 
	mutate(age = "total_pop")

pop90 <- pop90_fetch %>% 
	bind_rows(total90) %>% 
	ungroup() %>% 
	select(-state) %>%
	mutate(moe = 0) %>% 
	rename(var = age, estimate = value)
```

## Clean

```{r}
deci_pop <- deci_pops_bind %>% 
	separate(label, into = c("total", "gender", "age"), sep = "!!", fill = "right") %>% 
	clean_names() %>% 
	filter(!variable %in% c("P012002", "P012026")) %>% #remove total male, total female
	mutate(age = if_else(variable == "P012001", "total_pop", age),
				 moe = 0) %>% 
	select(-gender, -total, -variable, -state) %>% 
	rename(estimate = value)

acs_pop <- sex_by_age_bind %>%
	separate(label, into = c("total", "gender", "age"), sep = "!!", fill = "right") %>% 
	clean_names() %>%
	filter(!grepl("_002|_026", variable)) %>% #remove total, total male, total female
  mutate(age = if_else(variable == "B01001_001", "total_pop", age)) %>% 
	select(-gender, -total, -variable, -state) %>% 
	bind_rows(deci_pop) %>% 
	mutate(age = str_replace_all(age, " ", "_") %>% 
				 	str_to_lower()) %>% 
	mutate(age = as.factor(age) %>% 
				 	fct_relevel(., "under_5_years", "5_to_9_years") %>% 
				 	fct_collapse(.,
				 							 `15_to_19_years` = c("15_to_17_years", "18_and_19_years"),
				 							 `20_to_24_years` = c("20_years", "21_years", "22_to_24_years"),
				 							 `60_to_64_years` = c("60_and_61_years", "62_to_64_years"),
				 							 `65_to_69_years` = c("65_and_66_years", "67_to_69_years"))) %>% 
	group_by(year, level, geoid, name, county, var = age) %>% 
	summarise(estimate = sum(estimate),
						moe = moe_sum(moe = moe, estimate = estimate)) %>% 
	ungroup() %>% 
	group_by(year, level, geoid, name, county) %>% 
	mutate(moe = replace_na(moe, 0),
				 moe = round(moe, 0)) %>% 
	bind_rows(pop90) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate, moe = moe)

age_lut <- tibble(agegrp = c(99, seq(1:18)),
									age = unique(acs_pop$var)) %>% 
	add_row(agegrp = 0, age = "under_5_years")

period_lut <- tibble(
	estimate_date = c(
		"remove_april_2000",
		"remove_july_2000",
		seq(2001, 2009),
		"remove_april_2010",
		"remove_july_2010"),
	year = seq(1:13))

int_pop <- intercensal %>% 
	mutate(geoid = paste(state, county, sep = "")) %>% 
  select(geoid, name = ctyname, year, agegrp, estimate = tot_pop) %>% 
	left_join(age_lut, by = "agegrp") %>% 
	left_join(period_lut, by = "year") %>% 
	filter(!grepl("remove", estimate_date)) %>% 
	select(-year, -agegrp) %>% 
	rename(var = age, year = estimate_date) %>% 
	mutate(year = as.numeric(year),
				 level = "2_counties", county = NA, moe = 0,
				 level = as.factor(level),
				 var = as.factor(var) %>% 
				 	fct_relevel(., "under_5_years", "5_to_9_years")) %>% 
	group_by(year, level, geoid, name, county, var) %>% 
	summarise(estimate = sum(estimate), moe = sum(moe)) %>% 
	ungroup() %>% 
	group_by(year, level, geoid, name, county) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate, moe = moe)

int_ct <- int_pop %>% 
	ungroup() %>% 
	select(-geoid, -level, -name, -share, -sharemoe) %>% 
	group_by(year, county, var) %>% 
	summarise(estimate = sum(estimate), moe = sum(moe)) %>% 
	mutate(name = "Connecticut", level = "1_state", geoid = "09",
				 level = as.factor(level)) %>% 
	ungroup() %>% 
	group_by(year, level, geoid, name, county) %>% 
	calc_shares(group = var, denom = "total_pop", value = estimate, moe = moe)

pop_by_age_out <- bind_rows(int_pop, int_ct, acs_pop) %>% 
	mutate(level = fct_relevel(level, "1_state", "2_counties", "3_towns")) %>% 
	arrange(level, geoid, year)

pop_by_age_out %>% 
	write_csv(., "../output_data/pop_by_age_1990_2018.csv")
```

## Calculate change

```{r}
age_change <- pop_by_age_out %>%
	select(-share, -moe, -sharemoe) %>%
	rename(age = var) %>% 
	group_by(level, geoid, county, age) %>%
	arrange(name, year, age) %>%
	mutate(diff = estimate - lag(estimate, default = first(estimate))) %>%
	arrange(level, geoid, year, age) %>%
	mutate(measure = "pop_change_from_prev_data_year") %>%
	select(-estimate) %>%
	rename(estimate = diff) %>% 
	select(year, level, geoid, name, county, age, measure, estimate)
	
age_change %>% 
	write_csv("../output_data/pop_by_age_change_1990_2018.csv")
```

```{r}
pop_by_age_out %>% 
	ungroup() %>% 
	filter(level != "3_towns", year %in% c( 2000, 2010, 2018)) %>% 
	select(year, name, var, estimate) %>% 
	mutate(year = as.factor(year),
				 var = fct_collapse(var,
														`Under 19 years` = c("under_5_years", "5_to_9_years", "10_to_14_years", "15_to_19_years"),
														`20–34 years` = c("20_to_24_years", "25_to_29_years", "30_to_34_years"),
														`35–59 years` = c("35_to_39_years", "40_to_44_years", "45_to_49_years", "50_to_54_years", "55_to_59_years"),
														`60–74 years` = c("60_to_64_years", "65_to_69_years", "70_to_74_years"),
														`75 years and over` = c("75_to_79_years", "80_to_84_years", "85_years_and_over"))) %>% 
	group_by(year, name, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	group_by(year, name) %>% 
	calc_shares(group = var, denom = "total_pop") %>% 
	filter(var != "total_pop", name == "Connecticut") %>% 
	ggplot(aes(year, estimate, group = var)) +
	geom_vline(aes(xintercept = year), size = .5, color = "grey70") +
	geom_point(aes(color = var), size = 4) +
	geom_line(aes(color = var), size = 1) +
  geom_text(aes(label = scales::comma(estimate, accuracy = 1), vjust = -1.4, family = "Roboto Condensed")) +
	scale_y_continuous(expand = expansion(mult = c(.1, .125))) +
	hrbrthemes::theme_ipsum_rc() +
	guides(color = guide_legend(title = "", override.aes = list(linetype = 0))) +
	labs(title = "Population by age group, 2000–2018",
			 subtitle = "Connecticut",
			 x = "", y = "") +
	theme(plot.title.position = "plot",
				axis.text.y = element_blank(),
				panel.grid.major = element_blank(),
				panel.grid.minor = element_blank(),
				axis.text.x = element_text(colour = "black"))
```

```{r fig.height=6}
pop_by_age_out %>% 
	ungroup() %>% 
	filter(level != "3_towns", year %in% c( 2000,  2018)) %>% 
	select(year, name, var, estimate) %>% 
	mutate(year = as.factor(year),
				 var = fct_collapse(var,
				 									 `Total population` = "total_pop",
				 									 `Under 19 years` = c("under_5_years", "5_to_9_years", "10_to_14_years", "15_to_19_years"),
				 									 `20–34 years` = c("20_to_24_years", "25_to_29_years", "30_to_34_years"),
				 									 `35–59 years` = c("35_to_39_years", "40_to_44_years", "45_to_49_years", "50_to_54_years", "55_to_59_years"),
				 									 `60–74 years` = c("60_to_64_years", "65_to_69_years", "70_to_74_years"),
				 									 `75 years and over` = c("75_to_79_years", "80_to_84_years", "85_years_and_over"))) %>% 
	group_by(year, name, var) %>% 
	summarise(estimate = sum(estimate)) %>% 
	ungroup() %>% 
	pivot_wider(id_cols = c("name", "var"), names_from = year, values_from = estimate) %>% 
	mutate(pct_chg = (`2018` - `2000`) / `2000`,
				 lbl_chg = scales::percent(pct_chg, accuracy = 1),
				 `2000` = scales::comma(`2000`, accuracy = 1),
				 `2018` = scales::comma(`2018`, accuracy = 1)) %>% 
	select(Name = name, `Age group` = var, `Estimate, 2000` = `2000`, `Estimate, 2018` = `2018`, `Pct. chg. 2000–2018` = lbl_chg) %>% 
	kableExtra::kable()
```